---
title: "Examining Factors that Affect Student Performance"
author: "John Russell"
date: "2023-01-04"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The goal of this project is to gain insight on classroom performance. We will begin with a series of visualizations and conclude with a predictive model. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
library(plotly)
library(tidyverse)
library(gtExtras)
library(gt)
score_data <- read.csv('StudentsPerformance.csv', header = TRUE)
attach(score_data)

score_data <- mutate(score_data, mean.score =
                             rowMeans(select(
                                     score_data, c(math.score,
                                                   reading.score,
                                                   writing.score)
                             ), na.rm = TRUE))
```

### We will first examine the distribution of subject specific scores across gender.

```{r echo=FALSE, message=FALSE, warning=FALSE}
male_score_data <- filter(score_data, gender == 'male')

female_score_data <- filter(score_data, gender == 'female')

#### MATH SCORES
male_math <-
        ggplot(male_score_data, aes(x = math.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Math Score", y = "Count", title = 'Male Math') +
        geom_density(alpha = .2, fill = 'blue') +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_math <-
        ggplot(female_score_data, aes(x = math.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Math Score", y = "Count", title = 'Female Math') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




#### READING SCORES
male_reading <-
        ggplot(male_score_data, aes(x = reading.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Reading Score", y = "Count", title = 'Male Reading') +
        geom_density(alpha = .2, fill = 'blue')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_reading <-
        ggplot(female_score_data, aes(x = reading.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Reading Score", y = "Count", title = 'Female Reading') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




#### WRITING SCORES

male_writing <-
        ggplot(male_score_data, aes(x = writing.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Writing Score", y = "Count", title = 'Male Writing') +
        geom_density(alpha = .2, fill = 'blue')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_writing <-
        ggplot(female_score_data, aes(x = writing.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Writing Score", y = "Count", title = 'Female Writing') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




# PLOTS

plots_arranged <-
        (male_math | female_math) /
        (male_reading | female_reading) /
        (male_writing |
                 female_writing) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                                                   title = "Distribution of Scores across Gender")

plots_arranged
```

```{r echo=FALSE}
male_means <-
        c(
                "Male",
                round(mean(male_score_data$math.score), 2),
                round(mean(male_score_data$reading.score), 2),
                round(mean(male_score_data$writing.score),2)
        )

female_means <-
        c(
                "Female",
                round(mean(female_score_data$math.score),2),
                round(mean(female_score_data$reading.score),2),
                round(mean(female_score_data$writing.score),2)
        )


mean_matrix <- as.data.frame(rbind(male_means, female_means))
colnames(mean_matrix) <- c('Gender', 'Math', 'Reading', "Writing")
mean_df <- as.data.frame(mean_matrix)



mean_df %>% gt() %>% gt_theme_nytimes() %>% tab_header(title = 'Mean Scores')
```

\textbf{There are some interesting conclusion we can make:}

\begin{itemize}
\item \textbf{Male's tend to score higher in Math.}
\item \textbf{Female's tend to score higher in Reading and Writing.}
\item \textbf{The distribution for the male group is relatively normal while the distribution for the female group is skewed to the right.}
\end{itemize}


## Next, we want to look at the effect of parental level of education on subject scores.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

parents1 <- filter(score_data, parental.level.of.education == "some high school")
parents2 <- filter(score_data, parental.level.of.education == "high school")
parents3 <- filter(score_data, parental.level.of.education == "some college")
parents4 <- filter(score_data, parental.level.of.education == "associate's degree")
parents5 <- filter(score_data, parental.level.of.education == "bachelor's degree")
parents6 <- filter(score_data, parental.level.of.education == "master's degree")


parents <- score_data %>% count(parental.level.of.education)

list<- c("associate's", "bachelor's", "high school", "master's", 
         "some college", "some high school")

### For Loop to change row names

for (i in 1:6) {
        parents$parental.level.of.education[i] = list[i]
}

library(gtExtras)
parents <- parents %>% mutate(percent = (n/sum(n))*100)
parents %>% gt() %>% gt_theme_nytimes()



library(devtools)


colnames(parents) <- c('ParentalEducation', "n", "percent")

hi<-ggplot(parents, aes(x=ParentalEducation, y=n, fill=ParentalEducation)) + 
        geom_col() + 
        theme(legend.position = "none")+
        labs(x="", y = "Count", title = 'Parental Level of Education')

pieplot <- ggplot(parents, aes(x="", y=percent, fill=ParentalEducation)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
        labs(y="", x="")



```


```{r echo=FALSE, message=FALSE, warning=FALSE}
#Some level of college

parents_college <- filter(score_data, parental.level.of.education %in% c("associate's degree", "bachelor's degree", "master's degree", "some college"))

parents_nocollege <- filter(score_data, parental.level.of.education %in% c("high school", "some high school"))


f1<-ggplot(parents_college, aes(x=mean.score)) + geom_histogram(aes(y=..density..),fill='darkgreen', color='black') +
        scale_x_continuous(breaks = seq(0, 100, 25))+
        labs(title="Mean Score with Some College")+
        geom_density(alpha=.2, fill='green')

f2<-ggplot(parents_nocollege, aes(x=mean.score)) + geom_histogram(aes(y=..density..), fill="darkred", color='black') +
        scale_x_continuous(breaks = seq(0, 100, 25)) +
        labs(title="Mean Scores with No College")+
        geom_density(alpha=.2, fill='red')


plot <- hi/(f1|f2) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Distribution of Mean Scores & Some College vs. No College")
plot

```

\textbf{There are a few things to notice:}

\begin{itemize}
\item \textbf{Most students have parents who achieved some level of college ($62.5\%$).}
\item  \textbf{The median score for students whose parents achieved some level of college is $70.3\%$.}
\item  \textbf{The median score for kids whose parents did not reach college is $65.7\%$. }
\end{itemize}

```{r echo=FALSE}

p1<-ggplot(parents1, aes(x=mean.score)) + geom_histogram(fill= '#f564e3',color='black', binwidth = 2.5) + 
        labs(title="Some High School", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10)) +
        scale_y_continuous(NULL, breaks = NULL) 

p2<-ggplot(parents2, aes(x=mean.score)) + geom_histogram(fill='#00ba38',color='black', binwidth = 2) + 
        labs(title = "High School", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p3<-ggplot(parents3, aes(x=mean.score)) + geom_histogram(fill='#619cff',color='black', binwidth = 2) + 
        labs(title="Some College", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p4<-ggplot(parents4, aes(x=mean.score)) + geom_histogram(fill='#f8766d',color='black', binwidth = 1.7) + 
        labs(title = "Associate's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p5<-ggplot(parents5, aes(x=mean.score)) + geom_histogram(fill='#b79f00',color='black', binwidth = 2) + 
        labs(title="Bachelor's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p6<-ggplot(parents6, aes(x=mean.score)) + geom_histogram(fill='#00bfc4',color='black', binwidth = 2) + 
        labs(title = "Master's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)+
        geom_density(alpha = .2, fill = 'pink')

ed_arranged <- ((p1|p2) / (p3|p4) / (p5|p6)) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Distribution of Mean Scores across Parental Education Level") + theme_dark()
ed_arranged

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.height=4, fig.width=8}
histo_df <- data.frame(values <- c(parents1$mean.score, parents2$mean.score, parents3$mean.score,
                                   parents4$mean.score, parents5$mean.score, parents6$mean.score),
                       group = c(rep("some hig.", length(parents1$mean.score)),
                                 rep("high s", length(parents2$mean.score)),
                                 rep("some col.", length(parents3$mean.score)),
                                 rep("associate's", length(parents4$mean.score)),
                                 rep("bachelor's", length(parents5$mean.score)),
                                 rep("master's", length(parents6$mean.score))))

overlay_histo <- ggplot(histo_df, aes(x=values, fill=group)) + geom_density(position="identity", alpha=0.3) + 
        theme(legend.position = "bottom") + labs(x="Mean Score", y="")

overlay_histo+ plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Overlayed Density Plot of Mean Scores")

```

\textbf{We see that student's whose parents achieved beyond a bachelor's degree are more likely to score beyond $80\%$. On the other hand, student's whose parents did not attend college are more likely to score below $60\%$. Otherwise, it appears to be difficult to differentiate between different levels of parental education.}

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggpubr)
groupA <- filter(score_data, race.ethnicity == "group A")
a <-
        ggplot(
                count(groupA, parental.level.of.education),
                aes(x = parental.level.of.education, y = n, fill = parental.level.of.education)
        ) +
        geom_col() + theme(legend.position = "bottom", axis.text.x = element_blank()) + labs(title = "Group A", x =
                                                                                                     "Parental Education")

groupB <- filter(score_data, race.ethnicity == "group B")
b <-
        ggplot(
                count(groupB, parental.level.of.education),
                aes(x = parental.level.of.education, y = n, fill = parental.level.of.education)
        ) +
        geom_col() + theme(legend.position = "bottom", axis.text.x = element_blank()) +
        labs(title = "Group B", x = "Parental Education")

groupC <- filter(score_data, race.ethnicity == "group C")
c <-
        ggplot(
                count(groupC, parental.level.of.education),
                aes(x = parental.level.of.education, y = n, fill = parental.level.of.education)
        ) +
        geom_col() + theme(legend.position = "bottom", axis.text.x = element_blank()) +
        labs(title = "Group C", x = "Parental Education")

groupD <- filter(score_data, race.ethnicity == "group D")
d <-
        ggplot(
                count(groupD, parental.level.of.education),
                aes(x = parental.level.of.education, y = n, fill = parental.level.of.education)
        ) +
        geom_col() + theme(legend.position = "bottom", axis.text.x = element_blank()) +
        labs(title = "Group D", x = "Parental Education")

groupE <- filter(score_data, race.ethnicity == "group E")
e <-
        ggplot(
                count(groupE, parental.level.of.education),
                aes(x = parental.level.of.education, y = n, fill = parental.level.of.education)
        ) +
        geom_col() + theme(legend.position = "bottom", axis.text.x = element_blank()) +
        labs(title = "Group E", x = "Parental Education")


ggarrange(
        a,
        b,
        c,
        d,
        e,
        ncol = 2,
        nrow = 3,
        common.legend = TRUE,
        legend = "bottom"
)
ggplot(score_data, aes(x=mean.score, y=race.ethnicity)) + geom_count()

```


```{r message=FALSE, warning=FALSE, include=FALSE}

A_noCollege <- round(((
        count(groupA, parental.level.of.education)[6, 2] +
                count(groupA, parental.level.of.education)[3, 2]
) /
        count(score_data, race.ethnicity)[1, 2]) * 100,2)

B_noCollege <- round(((
        count(groupB, parental.level.of.education)[6, 2] +
                count(groupB, parental.level.of.education)[3, 2]
) /
        count(score_data, race.ethnicity)[2, 2]) * 100,2)

C_noCollege <- round(((
        count(groupC, parental.level.of.education)[6, 2] +
                count(groupC, parental.level.of.education)[3, 2]
) /
        count(score_data, race.ethnicity)[3, 2]) * 100,2)

D_noCollege <- round(((
        count(groupD, parental.level.of.education)[6, 2] +
                count(groupD, parental.level.of.education)[3, 2]
) /
        count(score_data, race.ethnicity)[4, 2]) * 100,2)

E_noCollege <- round(((
        count(groupE, parental.level.of.education)[6, 2] +
                count(groupE, parental.level.of.education)[3, 2]
) /
        count(score_data, race.ethnicity)[5, 2]) * 100, 2)

library(gt)

data.frame(
        A_noCollege,
        B_noCollege,
        C_noCollege,
        D_noCollege,
        E_noCollege) %>% gt() %>% gt_theme_nytimes()

```


## Test Preparation Courses

### The following charts compare the effect of a test preparation course on mean exam score. 

```{r echo=FALSE}

test_complete <- filter(score_data, test.preparation.course =="completed")
test_none <- filter(score_data, test.preparation.course =="none") 

c <-ggplot(test_complete, aes(x=mean.score)) + geom_histogram(binwidth = 3, aes(y=..density..), fill="darkblue", alpha=.4) + 
        geom_density(alpha=.2, fill='blue') + geom_vline(aes(xintercept=(mean(mean.score))), linetype="dashed") + 
        scale_x_continuous(breaks = seq(0, 100, 10)) + labs(title="Test Prep")

n <- ggplot(test_none, aes(x=mean.score)) + geom_histogram(binwidth = 3, aes(y=..density..), fill="darkred", alpha=.4) + 
        geom_density(alpha=.2, fill='pink') + geom_vline(aes(xintercept=mean(mean.score)), linetype="dashed") +
        scale_x_continuous(breaks = seq(0, 100, 10)) + labs(title= "No Test Prep")

(c/n) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Test Preparation vs. Mean Score")

Subject <- c("Math", "Reading", "Writing", "Mean")
Prep <- c(round(mean(test_complete$math.score),2), 
          round(mean(test_complete$reading.score),2),
          round(mean(test_complete$writing.score),2), 
          round(mean(test_complete$mean.score),2))

No_Prep <- c(round(mean(test_none$math.score),2), 
          round(mean(test_none$reading.score),2),
          round(mean(test_none$writing.score),2), 
          round(mean(test_none$mean.score),2))

test_df <- data.frame(Subject, Prep, No_Prep)
test_df %>% gt() %>% gt_theme_nytimes() %>% tab_header(title = 'Prep vs. No Prep Exam Scores')
```

We see that a single test preparation course has a significant effect on exam scores. In fact, taking part in a test preparation course leads to an $11.7\%$ increase in mean exam score. We will examine this further when we construct our model. 


## Lunch

### The next piece of our data we will examine is student lunches. The chart below compares two levels of lunches: free/reduced and standard.


```{r echo=FALSE}

library(gt)
library(gtExtras)

lunch_standard <- filter(score_data, lunch=="standard")

lunch_free <- filter(score_data, lunch=="free/reduced")



lunch_df <- data.frame(values <- c(lunch_standard$mean.score,
                                   lunch_free$mean.score),
                       group = c(rep(
                               "standard", length(lunch_standard$mean.score)
                       ),
                       rep(
                               "free/reduced", length(lunch_free$mean.score)
                       )))



lunch_histo <- ggplot(lunch_df, aes(x=values, fill=group)) + geom_density(position = "identity", alpha=.3) + 
        theme(legend.position = "bottom") + labs(x="Mean Score", title = "Density Plot for Student Lunch")  +
        scale_x_continuous(breaks = seq(0, 100, 10))


lunch_histo

Subject <- c("Math", "Reading", "Writing", "Mean")
Standard <- c(round(mean(lunch_standard$math.score),2), 
          round(mean(lunch_standard$reading.score),2),
          round(mean(lunch_standard$writing.score),2), 
          round(mean(lunch_standard$mean.score),2))

Free <- c(round(mean(lunch_free$math.score),2), 
          round(mean(lunch_free$reading.score),2),
          round(mean(lunch_free$writing.score),2), 
          round(mean(lunch_free$mean.score),2))

lunch_df <- data.frame(Subject, Standard, Free)
lunch_df %>% gt() %>% gt_theme_nytimes() %>% tab_header(title = 'Lunch Type vs. Exam Scores')
```





