---
title: "Predicting Student Performance"
author: "John Russell"
date: "2023-01-04"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The goal of this project is to gain insight on classroom performance. We will begin with a series of visualizations and conclude with a predictive model. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
library(plotly)
library(tidyverse)
library(gtExtras)
library(gt)
score_data <- read.csv('StudentsPerformance.csv', header = TRUE)
attach(score_data)

score_data <- mutate(score_data, mean.score =
                             rowMeans(select(
                                     score_data, c(math.score,
                                                   reading.score,
                                                   writing.score)
                             ), na.rm = TRUE))
```

### We will first examine the distribution of subject specific scores across gender.

```{r echo=FALSE, message=FALSE, warning=FALSE}
male_score_data <- filter(score_data, gender == 'male')

female_score_data <- filter(score_data, gender == 'female')

#### MATH SCORES
male_math <-
        ggplot(male_score_data, aes(x = math.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Math Score", y = "Count", title = 'Male Math') +
        geom_density(alpha = .2, fill = 'blue') +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_math <-
        ggplot(female_score_data, aes(x = math.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Math Score", y = "Count", title = 'Female Math') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




#### READING SCORES
male_reading <-
        ggplot(male_score_data, aes(x = reading.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Reading Score", y = "Count", title = 'Male Reading') +
        geom_density(alpha = .2, fill = 'blue')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_reading <-
        ggplot(female_score_data, aes(x = reading.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Reading Score", y = "Count", title = 'Female Reading') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




#### WRITING SCORES

male_writing <-
        ggplot(male_score_data, aes(x = writing.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'darkblue',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Writing Score", y = "Count", title = 'Male Writing') +
        geom_density(alpha = .2, fill = 'blue')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))

female_writing <-
        ggplot(female_score_data, aes(x = writing.score)) + geom_histogram(
                aes(y = ..density..),
                binwidth = 2,
                fill = 'red',
                color = 'black',
                alpha = .4
        ) +
        labs(x = "Writing Score", y = "Count", title = 'Female Writing') +
        geom_density(alpha = .2, fill = 'pink')  +
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(0, 100, 25))




# PLOTS

plots_arranged <-
        (male_math | female_math) /
        (male_reading | female_reading) /
        (male_writing |
                 female_writing) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                                                   title = "Distribution of Scores across Gender")

plots_arranged
```

```{r echo=FALSE}
male_means <-
        c(
                "Male",
                round(mean(male_score_data$math.score), 2),
                round(mean(male_score_data$reading.score), 2),
                round(mean(male_score_data$writing.score),2)
        )

female_means <-
        c(
                "Female",
                round(mean(female_score_data$math.score),2),
                round(mean(female_score_data$reading.score),2),
                round(mean(female_score_data$writing.score),2)
        )


mean_matrix <- as.data.frame(rbind(male_means, female_means))
colnames(mean_matrix) <- c('Gender', 'Math', 'Reading', "Writing")
mean_df <- as.data.frame(mean_matrix)



mean_df %>% gt() %>% gt_theme_538() %>% tab_header(title = 'Mean Scores')
```

### We can see that male's tend to score higher in Math while female's tend to score higher in Reading and Writing. We see that the Reading and Writing histograms for the female group are skewed to the right. The 'why' is an interesting question in and of itself.

### Next, we want to look at the effect of parental level of education on subject scores.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

parents1 <- filter(score_data, parental.level.of.education == "some high school")
parents2 <- filter(score_data, parental.level.of.education == "high school")
parents3 <- filter(score_data, parental.level.of.education == "some college")
parents4 <- filter(score_data, parental.level.of.education == "associate's degree")
parents5 <- filter(score_data, parental.level.of.education == "bachelor's degree")
parents6 <- filter(score_data, parental.level.of.education == "master's degree")


parents <- score_data %>% count(parental.level.of.education)

list<- c("associate's", "bachelor's", "high school", "master's", 
         "some college", "some high school")

### For Loop to change row names

for (i in 1:6) {
        parents$parental.level.of.education[i] = list[i]
}

library(gtExtras)
parents <- parents %>% mutate(percent = (n/sum(n))*100)
parents %>% gt() %>% gt_theme_nytimes()



library(devtools)


colnames(parents) <- c('ParentalEducation', "n", "percent")

hi<-ggplot(parents, aes(x=ParentalEducation, y=n, fill=ParentalEducation)) + 
        geom_col() + 
        theme(legend.position = "none")+
        labs(x="", y = "Count", title = 'Parental Level of Education')

pieplot <- ggplot(parents, aes(x="", y=percent, fill=ParentalEducation)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
        labs(y="", x="")



```


```{r echo=FALSE, message=FALSE, warning=FALSE}
#Some level of college

parents_college <- filter(score_data, parental.level.of.education %in% c("associate's degree", "bachelor's degree", "master's degree", "some college"))

parents_nocollege <- filter(score_data, parental.level.of.education %in% c("high school", "some high school"))


f1<-ggplot(parents_college, aes(x=mean.score)) + geom_histogram(aes(y=..density..),fill='darkgreen', color='black') +
        scale_x_continuous(breaks = seq(0, 100, 25))+
        labs(title="Mean Score with Some College")+
        geom_density(alpha=.2, fill='green')

f2<-ggplot(parents_nocollege, aes(x=mean.score)) + geom_histogram(aes(y=..density..), fill="darkred", color='black') +
        scale_x_continuous(breaks = seq(0, 100, 25)) +
        labs(title="Mean Scores with No College")+
        geom_density(alpha=.2, fill='red')


plot <- hi/(f1|f2) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Distribution of Mean Scores & Some College vs. No College")
plot

```

\textbf{There are a few things to notice:}

\begin{itemize}
\item \textbf{Most students have parents who achieved some level of college ($65.7\%$).}
\item  \textbf{The median score for students whose parents achieved some level of college is $70.3\%$.}
\item  \textbf{The median score for kids whose parents did not reach college is $65.7\%$. }
\end{itemize}

```{r echo=FALSE}

p1<-ggplot(parents1, aes(x=mean.score)) + geom_histogram(fill= '#f8766d',color='black', binwidth = 2.5) + 
        labs(title="Some High School", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10)) +
        scale_y_continuous(NULL, breaks = NULL) 

p2<-ggplot(parents2, aes(x=mean.score)) + geom_histogram(fill='#b79f00',color='black', binwidth = 2) + 
        labs(title = "High School", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p3<-ggplot(parents3, aes(x=mean.score)) + geom_histogram(fill='#00ba38',color='black', binwidth = 2) + 
        labs(title="Some College", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p4<-ggplot(parents4, aes(x=mean.score)) + geom_histogram(fill='#00bfc4',color='black', binwidth = 1.7) + 
        labs(title = "Associate's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p5<-ggplot(parents5, aes(x=mean.score)) + geom_histogram(fill='#619cff',color='black', binwidth = 2) + 
        labs(title="Bachelor's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)

p6<-ggplot(parents6, aes(x=mean.score)) + geom_histogram(fill='#f564e3',color='black', binwidth = 2) + 
        labs(title = "Master's Degree", x="Mean Score") +
        scale_x_continuous(breaks = seq(0, 100, 10))+
        scale_y_continuous(NULL, breaks = NULL)+
        geom_density(alpha = .2, fill = 'pink')

ed_arranged <- ((p1|p2) / (p3|p4) / (p5|p6)) + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Distribution of Mean Scores across Parental Education Level")
ed_arranged

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
histo_df <- data.frame(values <- c(parents1$mean.score, parents2$mean.score, parents3$mean.score,
                                   parents4$mean.score, parents5$mean.score, parents6$mean.score),
                       group = c(rep("some high school", length(parents1$mean.score)),
                                 rep("high school", length(parents2$mean.score)),
                                 rep("some college", length(parents3$mean.score)),
                                 rep("associate's degree", length(parents4$mean.score)),
                                 rep("bachelor's degree", length(parents5$mean.score)),
                                 rep("master's degree", length(parents6$mean.score))))

overlay_histo <- ggplot(histo_df, aes(x=values, fill=group)) + geom_density(position="identity", alpha=0.3) + 
        theme(legend.position = "bottom") + labs(x="Mean Score")

overlay_histo + plot_annotation(theme = theme_gray(base_family = 'mono'),
                    title = "Overlay Mean Scores across Parental Education Level")
```


